---
title: "2. Time series graphics"
author: "2.6 Scatterplots"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```

## Electricity Demand in Victoria, Australia

\fontsize{9}{9}\sf

```{r, echo=TRUE}
vic_elec_day_type <- vic_elec |> 
   filter(year(Time) == 2014) |> 
   mutate(Day_Type = case_when(
      Holiday ~ "Holiday",
      wday(Date) %in% 2:6 ~ "Weekday",
      TRUE ~ "Weekend"))
vic_elec_day_type
```

## Time plots

```{r, echo=FALSE, fig.height=3.6}
vic_elec_day_type |> 
   select(Temperature, Demand) |> 
   pivot_longer(-Time) |> 
   ggplot(aes(Time, value, colour = name)) +
   geom_line() +
   facet_grid(name ~ ., scales = "free_y") +
   guides(colour = "none") +
   labs(
    y = "Degrees Celsius                   GW         "
  )


```

## Scatterplots

```{r, echo=TRUE, fig.height=2.2, fig.width=6}
vic_elec_day_type |> 
  ggplot(aes(x = Temperature, y = Demand)) +
  geom_point() + 
   labs(x = "Temperature (degrees Celsius)", y = "Electricity demand (GW)")
```

## Scatterplots

```{r, echo=TRUE, fig.height=2.2, fig.width=6}
vic_elec_day_type |> 
  ggplot(aes(x = Temperature, y = Demand, colour = Day_Type)) +
  geom_point() + 
   labs(x = "Temperature (degrees Celsius)", y = "Electricity demand (GW)")
```

## Correlation coefficient

Measures the extent of a linear relationship between two variables ($y$ and $x$).
\vspace{0.4cm}

- $r=\frac{\sum_{t=1}^T(y_t-\bar{y})(x_t-\bar{x})}{\sqrt{\sum_{t=1}^T(y_t-\bar{y})^2} \sqrt{\sum_{t=1}^T(x_t-\bar{x})^2}}$

\begin{textblock}{7}(8,5.2)
\begin{itemize}
  \item Lies between -1 and 1
\end{itemize}
\end{textblock}


## US consumption expenditure
```{r, echo=FALSE, fig.height=5, fig.width=7.3}
us_change |>
   pivot_longer(-Quarter, names_to="Series") |>
   autoplot(value) +
   facet_grid(Series ~ ., scales = "free_y") + 
   labs(y = "% change")
```

## US consumption expenditure

\vspace{-0.1cm}

```{r, echo=TRUE, fig.height=5}
us_change |> GGally::ggpairs(columns = 2:6)
```

