---
title: "3. Time series decomposition"
author: "3.3 Moving averages"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
source("setup.R")

us_retail_employment <- us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)
```

## Moving averages

The simplest estimate of the trend-cycle uses **moving averages**.

\begin{block}{$m$-MA}
\[\hat{T}_t = \frac{1}{m} \sum_{j=-k}^k y_{t+j}\]
\end{block}
where $k=\frac{m-1}{2}$.

## Moving averages

\fontsize{10}{10}\sf

```{r ressales1, echo=TRUE, cache=TRUE}
global_economy |> filter(Country == "Australia") |>
  autoplot(Exports) + 
  labs(y="% of GDP", title= "Total Australian exports")
```

## Moving average smoothing

\fontsize{12}{12}\sf

```{r elecma, echo=FALSE, warning=FALSE, cache=TRUE}

format_num <- function(x) ifelse(is.na(x), "", format(x, nsmall = 2))
options(knitr.kable.NA = '')

aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `5-MA` = slider::slide_dbl(Exports, mean,.before = 2, .after = 2, .complete = TRUE)) 

out <- dplyr::bind_rows(
    head(aus_exports, 8),
    tail(aus_exports, 8)
  ) |>
  as_tibble() |>
  mutate_if(is.numeric, format_num) 

out[6,]=matrix(rep("...",3),nrow=1)

out <- dplyr::bind_rows(out[1:6,], 
                        out[11:16,]) |> 
  knitr::kable(booktabs=TRUE, digits=6)
#out <- kableExtra::kable_styling(out, latex_options = 'hold_position')
out
```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `3-MA` = slider::slide_dbl(Exports, mean,.before = 1, .after = 1, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`3-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 3-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `5-MA` = slider::slide_dbl(Exports, mean,.before = 2, .after = 2, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`5-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 5-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `7-MA` = slider::slide_dbl(Exports, mean,.before = 3, .after = 3, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`7-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 7-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `9-MA` = slider::slide_dbl(Exports, mean,.before = 4, .after = 4, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`9-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 9-MA") +
  guides(colour = guide_legend(title = "series")) 

```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `11-MA` = slider::slide_dbl(Exports, mean,.before = 5, .after = 5, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`11-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 11-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `13-MA` = slider::slide_dbl(Exports, mean,.before = 6, .after = 6, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`13-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 13-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing

\fontsize{10}{12}\sf

```{r, echo=FALSE, cache=TRUE, warning=FALSE,message=FALSE}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  transmute(Exports, `15-MA` = 
              slider::slide_dbl(Exports, mean,.before = 7, .after = 7, .complete = TRUE))

aus_exports |> 
  autoplot(Exports) +
  autolayer(aus_exports,`15-MA`, color = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports: 15-MA") +
  guides(colour = guide_legend(title = "series")) 
```

## Moving average smoothing
\fontsize{12}{14}\sf

So a moving average is an \orange{average of nearby points}

  *  observations nearby in time are also likely to be \orange{close in value}.
  *  average eliminates some \orange{randomness} in the data, leaving a smooth trend-cycle component.


\begin{center}\begin{tabular}{ll}
3-MA: & $\hat{T}_t = (y_{t-1} + y_{t} + y_{t+1})/3$\\[0.4cm]
5-MA: & $\hat{T}_t = (y_{t-2} + y_{t-1} + y_{t} + y_{t+1} + y_{t+2})/5$
\end{tabular}\end{center}

  *  each average computed by dropping \orange{oldest} observation and including  \orange{next} observation.
  *  averaging \orange{moves} through time series until trend-cycle computed at each observation possible.

## Endpoints
\fontsize{13}{15}\sf

\alert{Why is there no estimate at ends?}

  * For a 3 MA, there cannot be estimates at time 1 or time $T$ because the observations at time 0 and $T+1$ are not available.
  * Generally: there cannot be estimates at times near the endpoints.

\alert{The order of the MA}

  * larger order means smoother, flatter curve
  * larger order means more points lost at ends
  * \orange{order = length of season} \textit{or cycle} removes pattern
  * But so far odd orders?

## Centered MA

\fontsize{12}{13}\sf

**4 MA:**
$$ \frac14(y_{t-2}+y_{t-1}+y_t + y_{t+1})$$
$$\mbox{or}\qquad\hspace*{1.4cm}
\frac14(y_{t-1}+y_{t}+y_{t+1} + y_{t+2})$$

\pause

\vspace*{0.2cm}
**\orange{Solution:}** take a further 2-MA to "centre" result.
\vspace*{-0.2cm}

\begin{align*}
T_t &= \frac12\Big\{ \frac14(y_{t-2}+y_{t-1}+y_t + y_{t+1}) \\
& \mbox{} \hspace*{1.95cm} +  \frac14(y_{t-1}+y_{t}+y_{t+1} + y_{t+2})\Big\}\\[0.1cm]
 &=\frac18y_{t-2}+\frac14y_{t-1}+\frac14y_t + \frac14y_{t+1} +\frac18y_{t+2}
\end{align*}

## Centered MA
\fontsize{12}{13}\sf
\begin{tabular}{ccccc}
\toprule
Year&
Data&
4-MA&
$2\times4$-MA&
\\
\midrule
1992 Q1&
443.00&
\up{~}&
&
\\
1992 Q2&
410.00&
\up{451.25}&
&
\\
1992 Q3&
420.00&
\up{448.75}&
450.00&
\\
1992 Q4&
532.00&
\up{451.50}&
450.12&
\\
1993 Q1&
433.00&
\up{449.00}&
450.25&
\\
1993 Q2&
421.00&
\up{444.00}&
446.50&
\\
1993 Q3&
410.00&
\up{448.00}&
446.00&
\\
1993 Q4&
512.00&
\up{438.00}&
443.00&
\\
\vdots &
\vdots &
{\vdots}&
\vdots&
\\
\end{tabular}\vspace*{10cm}

## Moving average trend-cycle

A moving average of the same length as the season removes the seasonal
pattern.

* For quarterly data: use a $2\times4$ MA

* For monthly data: use a $2\times 12$ MA
$$\hat{T}_t = \frac{1}{24}y_{t-6} + \frac1{12}y_{t-5} + \cdots +
\frac{1}{12}y_{t+5} + \frac1{24}y_{t+6}$$

## Moving average trend-cycle
\fontsize{9}{11}\sf

```{r eval=FALSE, include=TRUE}
us_retail_employment_ma <- us_retail_employment |>
  mutate(
    `12-MA` = slider::slide_dbl(Employed, mean,
          .before = 5, .after = 6, .complete = TRUE),
    `2x12-MA` = slider::slide_dbl(`12-MA`, mean,
          .before = 1, .after = 0, .complete = TRUE)
  )

us_retail_employment_ma |> 
  autoplot(Employed, color = "gray") +
  autolayer(us_retail_employment_ma, vars(`2x12-MA`),
            color = "#D55E00") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```

## Moving average trend-cycle

```{r elecequip2, echo=FALSE, cache=TRUE, warning=FALSE}

us_retail_employment_ma <- us_retail_employment |>
  mutate(
    `12-MA` = slider::slide_dbl(Employed, mean,
          .before = 5, .after = 6, .complete = TRUE),
    `2x12-MA` = slider::slide_dbl(`12-MA`, mean,
          .before = 1, .after = 0, .complete = TRUE)
  )
autoplot(us_retail_employment_ma, Employed, color = "gray") +
  autolayer(us_retail_employment_ma, vars(`2x12-MA`),
            color = "#D55E00") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```
