---
title: "3. Time series decomposition"
author: "3.4 Classical decomposition "
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")

us_retail_employment <- us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)
```

## Trend-cycle
\fontsize{12}{14}\sf
\begin{block}{}
Additive decomposition: $y_t = T_t+S_t+R_t = \hat{T}_t + \hat{S}_t + \hat{R}_t$\\
Multiplicative decomposition: $y_t = T_t\times S_t\times R_t = \hat{T}_t\times \hat{S}_t\times\hat{R}_t$\end{block}
\pause


 * Estimate $\hat{T}$ using $(2\times m)$-MA if $m$ is even. Otherwise, estimate $\hat{T}$ using $m$-MA

Compute de-trended series

  * Additive decomposition: $y_t-\hat{T}_t$
  * Multiplicative decomposition: $y_t/\hat{T}_t$



**De-trending**

Remove smoothed series $\hat{T}_t$ from $y_t$ to leave $S_t$ and $R_t$.

  * Additive model: $y_t-\hat{T}_t  = (\hat{T}_t+\hat{S}_t+\hat{R}_t) - \hat{T}_t = \hat{S}_t+\hat{R}_t$
  * Multiplicative model: $\frac{y_t}{\hat{T}_t} = \frac{\hat{T}_t\times \hat{S}_t\times \hat{R}_t}{\hat{T}_t} = \hat{S}_t\times \hat{R}_t$

## Estimating seasonal component

  *  Seasonal index for each season is estimated as an \orange{average} of the detrended series for that season of successive  years.
  * E.g., take averages across all Januaries to get $S^{(1)}$ if your data is monthly. 
  * If necessary, adjust the seasonal indices so that:
    * for additive: $S^{(1)}+S^{(2)}+\ldots+S^{(12)}=0$ 
    * for multiplicative: $S^{(1)}+S^{(2)}+\ldots+S^{(12)}=m$
  * The seasonal component $\hat{S}_t$ simply consists of replications of the seasonal indices.


## Remainder component

Additive decomposition: $\hat{R}_t = y_t - \hat{T}_t - \hat{S}_t$

Multiplicative decomposition: $\hat{R}_t = y_t / (\hat{T}_t\hat{S}_t)$
\vspace{.5cm}
\pause

### Classical decomposition

  *  Choose additive or multiplicative depending on which gives the most stable components.
  *  For multiplicative model, this method of estimation is known as \orange{ratio-to-moving-average method}.

## US Retail Employment 

```{r echo=FALSE, warning=FALSE, fig.height=6, fig.width=12.5}
us_retail_employment |>
   model(classical_decomposition(Employed, type = "additive")) |>
   components() |>
   autoplot() + xlab("Year") +
   ggtitle("Classical additive decomposition of total US retail employment")
```

## US Retail Employment 

```{r echo=FALSE, warning=FALSE, fig.height=6, fig.width=12.5}
us_retail_employment |>
  model(classical_decomposition(Employed, type = "multiplicative")) |>
  components() |>
  autoplot() + xlab("Year") +
  ggtitle("Classical multiplicative decomposition of total US retail employment")
```

## US Retail Employment 
\fontsize{10}{12}\sf
```{r echo=TRUE, warning=FALSE, eval=FALSE}
us_retail_employment |>
  model(classical_decomposition(Employed, type = "additive")) |>
  components() |>
  autoplot() + xlab("Year") +
  ggtitle("Classical additive decomposition of total 
          US retail employment")
```


## Comments on classical decomposition

  *  Estimate of trend is \orange{unavailable} for first few and last few observations.
  *  \orange{Seasonal component repeats} from year to year. May not be realistic.
  *  \orange{Not robust to outliers}.
  *  Newer methods designed to overcome these problems.



