---
title: "5. The forecaster's toolbox"
author: "5.9 Evaluating distributional accuracy"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: yes
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
library(gganimate)
```

## Google closing stock prices

```{r}
google_stock <- gafa_stock |>
  filter(Symbol == "GOOG", year(Date) >= 2015) |> 
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)
google_2015 <- google_stock |> 
  filter(Symbol == "GOOG", year(Date) == 2015)
google_jan_2016 <- google_stock |>
  filter(Symbol == "GOOG", yearmonth(Date) == yearmonth("2016 Jan"))
google_fit <- google_2015 |>
  model(
    Mean = MEAN(Close),
    `Naïve` = NAIVE(Close),
    Drift = RW(Close ~ drift())
  )
google_fc <- google_fit |>
  forecast(google_jan_2016)
```

## Google closing stock prices

```{r, fig.height=2.8}
google_fc |>
  filter(.model == "Mean") |>
  autoplot(bind_rows(google_2015, google_jan_2016)) +
  labs(y = "$US", title = "Google closing stock prices: Mean forecasts") +
  guides(colour = guide_legend(title = "Forecast")) + ylim(490,880)
```


## Google closing stock prices

```{r, fig.height=2.8}
google_fc |>
  filter(.model == "Naïve") |>
  autoplot(bind_rows(google_2015, google_jan_2016)) +
  labs(y = "$US", title = "Google closing stock prices: Naïve forecasts") +
  guides(colour = guide_legend(title = "Forecast")) + ylim(490,880)
```


## Google closing stock prices

```{r, fig.height=2.8}
google_fc |>
  filter(.model == "Drift") |>
  autoplot(bind_rows(google_2015, google_jan_2016)) +
  labs(y = "$US", title = "Google closing stock prices: Drift forecasts") +
  guides(colour = guide_legend(title = "Forecast")) + ylim(490,880)
```

## Evaluating quantile forecasts

\begin{textblock}{9.5}(0.2,1.2)
\begin{alertblock}{}\vspace*{-0.2cm}
\begin{align*}
f_{p,t} &= \text{quantile forecast with prob. $p$ at time $t$.}\\
y_{t} &= \text{observation at time $t$}\\
& \text{Expect probability}(y_t < f_{p,t}) = p
\end{align*}
\end{alertblock}\vspace*{-0.3cm}
\uncover<2->{\begin{block}{Quantile score}\vspace*{-0.2cm}
$$
  Q_{p,t} = \begin{cases}
  2(1 - p) \big|y_t - f_{p,t}\big|, & \text{if $y_{t} < f_{p,t}$}\\
  2p \big|y_{t} - f_{p,t}\big|, & \text{if $y_{t} \ge f_{p,t}$} \end{cases}
$$
\end{block}}
\end{textblock}
\begin{textblock}{15}(0.2,6.4)
\uncover<4->{
\begin{itemize}\itemsep=0cm\parskip=0cm
\item Low $Q_{p,t}$ is good
\item Multiplier of 2 often omitted, but useful for interpretation
\item $Q_{p,t}$ like absolute error (weighted to account for likely exceedance)
\end{itemize}}
\end{textblock}
\begin{textblock}{6}(10,2)
\only<3->{\animategraphics[loop,autoplay]{10}{5-9-eval-distributions_files/figure-beamer/pinball-}{1}{100}}
\end{textblock}

```{r pinball, eval=FALSE, echo=FALSE, fig.show='animate', interval=1/10, message=FALSE, fig.height=3, fig.width=5, cache=FALSE}
# Turn eval=TRUE to recompute these graphs. They are loaded in the above animategraphics call.
prob <- seq(0.05, 0.95, by = 0.05)
df <- expand.grid(
  error = c(-10, 0, 10),
  p = c(prob, rev(head(prob, -1)[-1]))
) %>%
  mutate(
    state = rep(seq(length(p) / 3), rep(3, length(p) / 3)),
    qpt = 2 * p * error * (error > 0) - 2 * (1 - p) * error * (error < 0)
  )
labels <- df %>%
  select(p, state) %>%
  distinct() %>%
  mutate(label = paste0("p = ", sprintf("%.2f", p)))
df %>% ggplot(aes(x = error, y = qpt)) +
  geom_line(aes(group = state), colour = "red") +
  labs(
    x = latex2exp::TeX("Error: $y_t - f_{p,t}$"),
    y = latex2exp::TeX("Q_{p,t}")
  ) +
  geom_label(data = labels, aes(x = 0, y = 17, label = label)) +
  transition_states(state)
```

## Quantile scores
\fontsize{10}{10}\sf

```{r googlepcscore}
google_fc |>
  filter(.model == "Naïve", Date == "2016-01-04") |>
  accuracy(google_stock, list(qs=quantile_score), probs=0.1)
google_fc |>
  filter(.model == "Naïve", Date == "2016-01-04") |>
  accuracy(google_stock, list(qs=quantile_score), probs=0.9)
```

## Winkler Score

For $100(1-\alpha)$% prediction interval: $[\ell_{\alpha,t}, u_{\alpha,t}]$.\vspace*{-0.1cm}
$$
 W_{\alpha,t} = \frac{Q_{\alpha/2,t} + Q_{1-\alpha/2,t}}{\alpha}
   = \begin{cases}
  (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (\ell_{\alpha,t} - y_t) & \text{if } y_t < \ell_{\alpha,t} \\
  (u_{\alpha,t} - \ell_{\alpha,t})   & \text{if }  \ell_{\alpha,t} \le y_t \le u_{\alpha,t} \\
  (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (y_t - u_{\alpha,t}) & \text{if } y_t > u_{\alpha,t}.
  \end{cases}
$$

\fontsize{10}{10}\sf

```{r googlewinklerscore, dependson='GSPfc'}
google_fc |>
  filter(.model == "Naïve", Date == "2016-01-04") |>
  accuracy(google_stock, list(winkler = winkler_score), level = 80)
```

## Continuous Ranked Probability Score {-}

Average quantile scores over all values of $p$ to obtain the

**Continuous Ranked Probability Score** or CRPS.

\fontsize{10}{10}\sf

```{r crps, dependson='GSPfc'}
google_fc |>
  accuracy(google_stock, list(crps = CRPS))
```


## Scale-free comparisons using skill scores {-}

Skill scores provide a forecast accuracy measure relative to some benchmark method (often the naïve method).\vspace*{0.1cm}
$$
\text{CRPS\_SS}_{\text{Method}} = 
  \frac{\text{CRPS}_{\text{Naïve}} - \text{CRPS}_{\text{Method}}}{\text{CRPS}_{\text{Naïve}}}.
$$

\fontsize{10}{10}\sf

```{r skillscore, dependson='GSPfc'}
google_fc |>
  accuracy(google_stock, list(skill = skill_score(CRPS)))
```

