---
title: "7. Time series regression models"
author: "7.1 The linear model"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```

## Multiple regression and forecasting

\begin{block}{}\vspace*{-0.3cm}
\[
  y_t = \beta_0 + \beta_1 x_{1,t} + \beta_2 x_{2,t} + \cdots + \beta_kx_{k,t} + \varepsilon_t.
\]
\end{block}

* $y_t$ is the variable we want to predict: the "response" variable
* Each $x_{j,t}$ is numerical and is called a "predictor".
 They are usually assumed to be known for all past and future times.
* The coefficients $\beta_1,\dots,\beta_k$ measure the effect of each
predictor after taking account of the effect of all other predictors
in the model.

That is, the coefficients measure the \orange{marginal effects}.

* $\varepsilon_t$ is a white noise error term


## Example: US consumption expenditure
\fontsize{10}{11}\sf

```{r ConsInc, echo=FALSE, fig.width=10, fig.height=5, out.height="92%"}
us_change %>%
  pivot_longer(c(Consumption, Income), names_to="Series") %>%
  autoplot(value) +
  labs(y="% change")
```

## Example: US consumption expenditure
\fontsize{10}{11}\sf

```{r ConsInc2, echo=FALSE, fig.width=10, fig.height=5, out.height="92%"}
us_change %>% ggplot(aes(x=Income, y=Consumption)) +
  labs(y = "Consumption (quarterly % change)",
       x = "Income (quarterly % change)") +
  geom_point() + geom_smooth(method="lm", se=FALSE)
```

## Example: US consumption expenditure

\fontsize{7.8}{7.8}\sf
```{r, echo=TRUE}
fit_cons <- us_change %>%
  model(lm = TSLM(Consumption ~ Income))
report(fit_cons)
```


## Example: US consumption expenditure

```{r MultiPredictors, echo=FALSE, fig.height=5, fig.width=10, out.height="92%"}
us_change |>
  gather("Measure", "Change", Consumption, Income, Production, Savings, Unemployment) |>
  ggplot(aes(x = Quarter, y = Change, colour = Measure)) +
  geom_line() +
  facet_grid(vars(Measure), scales = "free_y") +
  labs(y = "") +
  guides(colour = "none")
```

## Example: US consumption expenditure

```{r ScatterMatrix, echo=FALSE, fig.height=5, fig.width=10, out.height="90%"}
us_change |>
  as_tibble() |>
  select(-Quarter) |>
  GGally::ggpairs()
```

##  Assumptions for the linear model
For forecasting purposes, we require the following assumptions:

* $\varepsilon_t$ have mean zero and are uncorrelated.

* $\varepsilon_t$ are uncorrelated with each $x_{j,t}$.
\pause

It is \orange{useful} to also have $\varepsilon_t \sim \text{N}(0,\sigma^2)$ when producing prediction intervals or doing statistical tests.

