---
title: "7. Time series regression models"
author: "7.1 The linear model"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```


## Least squares estimation

* In practice we need to estimate the coefficients: $\beta_0,\beta_1, \dots, \beta_k$.

$$\sum_{t=1}^T \varepsilon_t^2 = \sum_{t=1}^T (y_t -
  \beta_{0} - \beta_{1} x_{1,t} - \beta_{2} x_{2,t} - \cdots - \beta_{k} x_{k,t})^2$$ \pause

\vspace{.8cm}

```r
model(TSLM(y ~ x_1 + x_2 + ... + x_k))
```
* Estimated coefficients: $\hat\beta_0, \dots, \hat\beta_k$ 


## Example: US consumption expenditure

\fontsize{7}{6.8}\sf

```{r usestim, echo=TRUE}
fit_consMR <- us_change |>
  model(lm = TSLM(Consumption ~ Income + Production + Unemployment + Savings))
report(fit_consMR)
```

## Fitted values

\vspace{-0.5cm}

$$\hat{y}_t = \hat\beta_{0} + \hat\beta_{1} x_{1,t} + \hat\beta_{2} x_{2,t} + \cdots + \hat\beta_{k} x_{k,t}$$ \pause

```{r usfitted1, echo=FALSE}
augment(fit_consMR) |>
  ggplot(aes(x = Quarter)) +
  geom_line(aes(y = Consumption, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) +
  labs(y = NULL, title = "Percent change in US consumption expenditure") +
  scale_colour_manual(values = c(Data = "black", Fitted = "#D55E00")) +
  guides(colour = guide_legend(title = NULL))
```

## Example: US consumption expenditure

```{r usfitted2, echo=FALSE, message=FALSE, warning=FALSE}
augment(fit_consMR) |>
  ggplot(aes(y = .fitted, x = Consumption)) +
  geom_point() +
  labs(
    y = "Fitted (predicted values)",
    x = "Data (actual values)",
    title = "Percentage change in US consumption expenditure"
  ) +
  geom_abline(intercept = 0, slope = 1)
```

## Goodness of fit 

\begin{block}{Coefficient of determination }
\[
R^2 = \frac{\sum(\hat{y}_{t} - \bar{y})^2}{\sum(y_{t}-\bar{y})^2}
\]
\end{block}

\begin{block}{Standard error of the regression}
\[
  \hat{\sigma}_e=\sqrt{\frac{1}{T-k-1}\sum_{t=1}^{T}{e_t^2}}
\] where $k$ is the number of predictors in the model.
\end{block}
 

