---
title: "7. Time series regression models"
author: "7.1 Evaluating the regression model"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```



```{r usestim, include=FALSE}
fit_consMR <- us_change |>
  model(lm = TSLM(Consumption ~ Income + Production + Unemployment + Savings))
report(fit_consMR)
```

## Regression residuals

Residuals are defined as: 
\begin{align*}
  e_t &= y_t - \hat{y}_t \\
      &= y_t - \hat\beta_{0} - \hat\beta_{1} x_{1,t} - \hat\beta_{2} x_{2,t} - \cdots - \hat\beta_{k} x_{k,t}
\end{align*} \pause

Useful properties
\[
\sum_{t=1}^{T}{e_t}=0 \quad\text{and}\quad \sum_{t=1}^{T}{x_{k,t}e_t}=0\qquad\text{for all $k$}.
\]

## Checking assumptions

- $\varepsilon_t$ have mean zero and are uncorrelated, NID(0,$\sigma^2$).
- $\varepsilon_t$ are uncorrelated with each $x_{j,t}$.

   - Timeplot, ACF, Histogram (`gg_tsresiduals()`)
   - Against predictors (non-linearity)
   - Against fitted values (heteroscedasticity)
   - Against predictors not in the model (include predictor in the model)

Expect to see scatterplots resembling a horizontal band with no values too far from the band and no patterns such as curvature or increasing spread.


## Example: US consumption expenditure

```{r, echo=TRUE}
fit_consMR |> gg_tsresiduals()
```

## Example: US consumption expenditure

```{r, echo=FALSE, fig.width=10, fig.height=5, out.height="92%"}
us_change |>
  left_join(residuals(fit_consMR), by = "Quarter") |>
  pivot_longer(Income:Unemployment,
               names_to = "regressor", values_to = "x") |>
  ggplot(aes(x = x, y = .resid)) +
  geom_point() +
  facet_wrap(. ~ regressor, scales = "free_x") +
  labs(y = "Residuals", x = "")
```

## Example: US consumption expenditure

```{r, echo=FALSE, fig.width=10, fig.height=5, out.height="92%"}

augment(fit_consMR) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() + labs(x = "Fitted", y = "Residuals")
```