---
title: "7. Time series regression models"
author: "7.7 Nonlinear regression"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```


## Nonlinear regression

A \orange{log-log} functional form 

$$\log y=\beta_0+\beta_1 \log x +\varepsilon$$

where $\beta_1$ is interpreted as an elasticity (the average percentage change in $y$ resulting from a $1\%$ increase in $x$). \pause

- alternative specifications: log-linear, linear-log.
- use $\log(x+1)$ if required.
 

## Piecewise linear and regression splines

\fontsize{11}{11}\sf

\vspace{-0.6cm}

$$y=f(x) +\varepsilon$$ 
\vspace{-0.8cm}

where $f$ is a non-linear function. \pause

- For \orange{piecewise linear} let $x_1=x$ and 

\vspace{-0.8cm}

\begin{align*}
  x_{2} = (x-c)_+ &= \left\{
             \begin{array}{ll}
               0 & \text{if } x < c\\
               x-c &  \text{if } x \ge c.
             \end{array}\right.
\end{align*} \pause

\vspace{-0.6cm}

- In general, \orange{linear regression splines} 

\vspace{-0.6cm}

$$x_1=x~~~x_2=(x-c_1)_+~~~\ldots~~~x_k=(x-c_{k-1})_+$$  
where $c_1,\ldots,c_{k-1}$ are knots. \pause

* Need to select knots: can be difficult and arbitrary. 
* Automatic knot selection algorithms very slow.
* Using piecewise cubics achieves a smoother result.

Warning: better fit but forecasting outside the range of the historical data is even more unreliable.

## Nonlinear trends

\vspace{-0.6cm}

Piecewise linear trend with bend at $\tau$

\vspace*{-1.4cm}

\begin{align*}
x_{1,t} &= t \\
x_{2,t} &= \left\{ \begin{array}{ll}
  0 & t <\tau\\
  (t-\tau) & t \ge \tau
\end{array}\right.
\end{align*}

\pause

\vspace*{0.4cm}

Quadratic or higher order trend
\vspace*{-0.6cm}

\[
  x_{1,t} =t,\quad x_{2,t}=t^2,\quad \dots
\] \pause

\vspace*{-0.9cm}

\centerline{\alert{\textbf{NOT RECOMMENDED!}}}


## Example: Boston marathon winning times

```{r, fig.height=2.6, echo=TRUE}
marathon <- boston_marathon |>
  filter(Event == "Men's open division") |>
  select(-Event) |>
  mutate(Minutes = as.numeric(Time) / 60)
marathon |> autoplot(Minutes) + labs(y = "Winning times in minutes")
```

## Example: Boston marathon winning times

\fontsize{9}{9}\sf

```{r, echo=TRUE}
fit_trends <- marathon |>
  model(
    # Linear trend
    linear = TSLM(Minutes ~ trend()),
    # Exponential trend
    exponential = TSLM(log(Minutes) ~ trend()),
    # Piecewise linear trend
    piecewise = TSLM(Minutes ~ trend(knots = c(1940, 1980)))
  )
```

```{r}
fit_trends
```

## Example: Boston marathon winning times

```{r, echo=TRUE, eval=FALSE, fig.height=2.7}
fit_trends |>
  forecast(h = 10) |>
  autoplot(marathon)
```

```{r, echo=FALSE, message=TRUE, warning=FALSE, fig.height=3}
fc_trends <- fit_trends |> forecast(h = 10)
marathon |>
  autoplot(Minutes) +
  geom_line(
    data = fitted(fit_trends),
    aes(y = .fitted, colour = .model)
  ) +
  autolayer(fc_trends, alpha = 0.5, level = 95) +
  labs(
    y = "Minutes",
    title = "Boston marathon winning times"
  )
```

## Example: Boston marathon winning times

```{r residPiecewise, message=FALSE, warning=FALSE}
fit_trends |>
  select(piecewise) |>
  gg_tsresiduals()
```
