---
title: "9. ARIMA models"
author: "9.1 Unit root tests"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 2.8
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
library(patchwork)
```

## Unit root tests

Statistical tests to determine the required order of differencing.

  1. Augmented Dickey Fuller test: null hypothesis is that the data are \orange{non-stationary} and non-seasonal.
  2. Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test: null hypothesis is that the data are \orange{stationary} and non-seasonal.
  3. Other tests available for seasonal data.

\only<2>{\begin{textblock}{4}(10.2,4)
\begin{alertblock}{}
H$_0$: non-stationary
\end{alertblock}
\end{textblock}}

\only<2>{\begin{textblock}{3}(12,5.4)
\begin{alertblock}{}
H$_0$: stationary
\end{alertblock}
\end{textblock}}


## Example: Google stock price

```{r include=FALSE}
google_2018 <- gafa_stock |>
  filter(Symbol == "GOOG", year(Date) == 2018)
```


```{r}
google_2018 |>
  autoplot(Close) +
  labs(y = "Closing stock price ($USD)")
```

## Example: Google stock price

```{r}
google_2018 |>
  autoplot(difference(Close)) +
  labs(y = "Change in Google closing stock price ($USD)")
```

## KPSS test
\fontsize{9}{10}\sf



```{r, echo=TRUE}
google_2018 %>%
   features(Close, unitroot_kpss)
```

\pause

```{r, echo=TRUE}
google_2018 %>%
  features(Close, unitroot_ndiffs)
```


## Corticosteroid drug sales


```{r include=FALSE}
h02 <- PBS |>
  filter(ATC2 == "H02") |>
  summarise(Cost = sum(Cost) / 1e6)
```


```{r, echo=TRUE}
h02 |> autoplot(
  Cost
)
```

## Corticosteroid drug sales

```{r, echo=TRUE}
h02 |> autoplot(
  log(Cost)
)
```

## Corticosteroid drug sales

```{r, echo=TRUE}
h02 |> autoplot(
  log(Cost) |> difference(12)
)
```

## Corticosteroid drug sales

```{r, echo=TRUE}
h02 |> autoplot(
  log(Cost) |> difference(12) |> difference(1)
)
```

## Automatically selecting differences

\fontsize{13}{14}\sf

STL decomposition: $y_t = T_t+S_t+R_t$

Seasonal strength $F_s = \max\big(0, 1-\frac{\text{Var}(R_t)}{\text{Var}(S_t+R_t)}\big)$

If $F_s > 0.64$, do one seasonal difference.

\fontsize{10}{11}\sf

```{r, echo=TRUE}
h02 %>% mutate(log_sales = log(Cost)) %>%
  features(log_sales, feat_stl)
```

## Automatically selecting differences
\fontsize{10}{11}\sf

```{r, echo=TRUE}
h02 %>% mutate(log_sales = log(Cost)) %>%
  features(log_sales, unitroot_nsdiffs)
h02 %>% mutate(d_log_sales = difference(log(Cost), 12)) %>%
  features(d_log_sales, unitroot_ndiffs)
```