---
title: "9. ARIMA models"
author: "9.1 Random walk model"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 2.8
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
library(patchwork)
```

## Example: Google stock price

```{r}
google_2018 <- gafa_stock |>
  filter(Symbol == "GOOG", year(Date) == 2018)
```

## Example: Google stock price

```{r}
google_2018 |>
  autoplot(Close) +
  labs(y = "Closing stock price ($USD)")
```

## Example: Google stock price

```{r}
google_2018 |>
  ACF(Close) |>
  autoplot()
```

## Example: Google stock price

```{r}
google_2018 |>
  autoplot(difference(Close)) +
  labs(y = "Change in Google closing stock price ($USD)")
```

## Example: Google stock price

```{r}
google_2018 |>
  ACF(difference(Close)) |>
  autoplot()
```


## Example: Google stock price 

* The differences are the \orange{day-to-day} changes.
* Now the series looks just like a white noise series:
    * No autocorrelations outside the 95% limits.
* \orange{Conclusion:} The daily change in the Google stock price is essentially a random amount uncorrelated with previous days.


## Random walk model

If differenced series is white noise with zero mean:

\begin{block}{}
\centerline{$y_t-y_{t-1}=\varepsilon_t$ \hspace{0.4cm} or \hspace{0.4cm} $y_t=y_{t-1}+\varepsilon_t$}
\end{block}\vspace*{-0.3cm}
where $\varepsilon_t \sim NID(0,\sigma^2)$.

* Very widely used for non-stationary data.
* This is the model behind the \orange{naïve method}.\pause
* Random walks typically have:
    * long periods of apparent trends up or down
    * Sudden/unpredictable changes in direction \pause
* Forecast are equal to the last observation (naïve)
    * future movements up or down are equally likely.

## Random walk with drift model

If differenced series is white noise with non-zero mean:

\begin{block}{}
\centerline{$y_t-y_{t-1}=c+\varepsilon_t$ \hspace{0.4cm} or \hspace{0.4cm} $y_t=c+y_{t-1}+\varepsilon_t$}
\end{block}\vspace*{-0.3cm}
where $\varepsilon_t \sim NID(0,\sigma^2)$.

* $c$ is the \orange{average change} between consecutive observations.
* If $c>0$, $y_t$ will tend to drift upwards and vice versa.
* This is the model behind the \orange{drift method}.

## Seasonal differencing

If seasonally differenced data is white noise it implies: 
\begin{block}{}
\centerline{$y_t-y_{t-m}=\varepsilon_t$ \hspace{0.4cm} or \hspace{0.4cm} $y_t=y_{t-m}+\varepsilon_t$}
\end{block}
where $\varepsilon_t \sim NID(0,\sigma^2)$.

  * The model behind the \orange{seasonal naïve} method.
