---
title: "9. ARIMA models"
author: "9.3 Autoregressive models"
date: "OTexts.org/fpp3/"
classoption: aspectratio=169
titlepage: fpp3title.png
titlecolor: fpp3red
toc: false
output:
  binb::monash:
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    keep_tex: no
    includes:
      in_header: fpp3header.tex
---

```{r setup, include=FALSE}
source("setup.R")
library(patchwork)
```

## Autoregressive models

\fontsize{14}{15}\sf

\begin{block}{Autoregressive model - AR($p$):}
$$  y_{t} = c + \phi_{1}y_{t - 1} + \phi_{2}y_{t - 2} + \cdots + \phi_{p}y_{t - p} + \varepsilon_{t}, $$
\vspace{-0.2cm}
where $\varepsilon_t$ is white noise. This is a multiple regression with \orange{lagged values} of $y_t$ as predictors.
\end{block}

```{r arp, echo=FALSE, fig.height=2.4}
set.seed(1)
p1 <- tsibble(idx = seq_len(100), sim = 10 + arima.sim(list(ar = -0.8), n = 100), index = idx) %>%
  autoplot(sim) + labs(y="", title  = "AR(1)")
p2 <- tsibble(idx = seq_len(100), sim = 20 + arima.sim(list(ar = c(1.3, -0.7)), n = 100), index = idx) %>%
  autoplot(sim) + labs(y="", title ="AR(2)")
p1 | p2
```

## AR(1) model

\begin{block}{}
\centerline{$y_{t} = -0.8 y_{t - 1} + \varepsilon_{t}$}
\end{block}
\rightline{$\varepsilon_t\sim N(0,1)$,\quad $T=100$.}

```{r, echo=FALSE}
tsibble(idx = seq_len(100), sim = arima.sim(list(ar = -0.8), n = 100), index = idx) %>%
  autoplot(sim) + ylab("") + ggtitle("AR(1)")
```
			   
## AR(1) model

\begin{block}{}
\centerline{$y_{t}    = \phi_1 y_{t - 1}  +  \varepsilon_{t}$}
\end{block}

* When $\phi_1=0$, $y_t$ is \orange{equivalent to a WN}
* When $\phi_1=1$, $y_t$ is \orange{equivalent to a RW}
* We require $|\phi_1|<1$ for stationarity. The closer $\phi_1$ is to the bounds the more the process wanders above or below it's unconditional mean (zero in this case).
* When $\phi_1<0$, $y_t$ tends to \orange{oscillate between positive and negative values}.

\vspace{1cm}

## AR(1) model including a constant

\begin{block}{}
\centerline{$y_{t} = 18-0.8 y_{t - 1} + \varepsilon_{t}$}
\end{block}
\rightline{$\varepsilon_t\sim N(0,1)$,\quad $T=100$.}

```{r, echo=FALSE}
tsibble(idx = seq_len(100), sim = 10+ arima.sim(list(ar = -0.8), n = 100), index = idx) %>%
  autoplot(sim) + ylab("") + ggtitle("AR(1)")
```

## AR(1) model including a constant

\begin{block}{}
  \centerline{$y_{t} = c + \phi_1 y_{t - 1} + \varepsilon_{t}$}
\end{block}

* When $\phi_1=0$ and $c=0$, $y_t$ is \orange{equivalent to WN};
* When $\phi_1=1$ and $c=0$, $y_t$ is \orange{equivalent to a RW};
* When $\phi_1=1$ and $c\ne0$, $y_t$ is \orange{equivalent to a RW with drift};

\vspace{4cm}


## AR(1) model including a constant

\begin{block}{}
\centerline{$y_{t}    = c+\phi_1 y_{t - 1}  +  \varepsilon_{t}$}
\end{block}

* $c$ is related to the mean of $y_{t}$. 
* Let $E(y_{t})=\mu$ then $c=\mu\times(1-\phi_1)$.

\vspace{3cm}


* `ARIMA()` takes care of whether you need a constant or not, or you can overide it. 



## AR(1) model including a constant

* If included estimated model returns `w/ mean`

\fontsize{12}{12}\sf

```
Series: sim 
Model: ARIMA(1,0,0) w/ mean 

Coefficients:
          ar1  constant
      -0.8381   18.3527
s.e.   0.0540    0.1048

sigma^2 estimated as 1.11:  log likelihood=-146.7
AIC=299.4   AICc=299.7   BIC=307.2
```

## AR(2) model

\begin{block}{}
  \centerline{$y_t = 8 + 1.3y_{t-1} - 0.7 y_{t-2} + \varepsilon_t$}
\end{block}
\rightline{$\varepsilon_t\sim N(0,1)$, \qquad $T=100$.}

```{r echo=FALSE}
p2
```

## Stationarity conditions
\fontsize{14}{15}\sf

We normally restrict autoregressive models to stationary data, and then some constraints on the values of the parameters are required.

\begin{block}{General condition for stationarity}
  Complex roots of $1-\phi_1 z - \phi_2 z^2 - \dots - \phi_pz^p$ lie outside the unit circle on the complex plane.
\end{block}\pause

* For $p=1$: $-1<\phi_1<1$.
* For $p=2$:\newline $-1<\phi_2<1\qquad \phi_2+\phi_1 < 1 \qquad \phi_2 -\phi_1 < 1$.
* More complicated conditions hold for $p\ge3$.
* Estimation software takes care of this.

